Simple position restraint schemes for alchemical binding free energy calculations

### 2022-01-28 manuscript/main.tex
Updates from Overleaf

\begin{abstract}
Computed free energies have become particularly valuable in the process of hit discovery and ligand screening for novel therapeutics. Simulation techniques based on free energy perturbation require sampling of alchemical intermediates which de-couple the non-bonded interactions of a molecule that is typically bound to a receptor. Harmonic restraints are typically imposed In order to maintain a bound-state orientation while the ligand is de-coupled. These restraints fall into several categories. Center-of-mass (COM) restraints are sufficient for maintaining the ligand close to its original conformation while allowing for sampling multiple binding modes, but do not . Boresch restraints are commonly used to maintain a specific bound orientation, requiring the combined implementation of restrained bond lengths, angles, and dihedrals. In this chapter, we propose an intermediate approach that loosely maintains the bound orientation of a molecule using three position restraints as well as a simple analytical correction for the bias imposed by it.
\end{abstract}

%\pacs{Valid PACS appear here}% PACS, the Physics and Astronomy
                             % Classification Scheme.
\keywords{alchemical free energy perturbation, flat-histogram methods, ligand binding, Monte Carlo}%Use showkeys class option if keyword
                              %display desired
\maketitle

@@

\section*{Introduction}

A restrained double-decoupling approach has long been a standard procedure for calculating absolute binding free energies of protein-ligand binding.  In this approach, the free energy of alchemically decoupling the ligand is usually calculated in two steps: first, the free energy of \textit{restraining} the ligand is calculated, and then the free energy of decoupling the restrained ligand's nonbonded interactions is calculated.     For a ligand in complex with its protein receptor, the restraint is vitally important to keep the ligand from dissociating.  For a ligand in solution, the restraint free energy can be estimated analytically.

Currently, the most popular restraint scheme is a method by Boresch et al \cite{boresch2003absolute} that utilizes a set of orthogonal positional, angular, and dihedral restraints.  The advantage of this scheme is that it is an exact analytical expression for the free energy of restraint for a ligand in solution.    There are several disadvantages to this scheme, however.  One disadvantage is the practical difficulty of implementing three different kinds of restraints, and judiciously choosing the optimal degrees of freedom in which to impose them.  This has recently been emphasized by Proccaci et al. \cite{procacci2021ns} who show that large biases in free energies estimated using a non-equilibrium work (NEW) method can be introduced when the Boresch-restrained ligand pose does not match well with the actual pose, which is often unknown ahead of time.

In this work, we propose a simpler alternative to Boresch-style orientation restraints that uses three harmonic position restraints.  This scheme is easier to implement in practice, and is accurate when sufficiently rigid molecular substructures are restrained.  We derive an analytical expression for the free energy of restraint, and test the accuracy of the expression using a series of model compounds with various combinrations of rigid aromatic rings and flexible aliphatic chains.   These "Hurley-style" restraints may be useful in many applications due to their simplicity.

@@

In practice, bottlenecks often arise in the Monte Carlo sampling that lead to poor convergence of the free energy estimates.  One source of sampling bottlenecks comes from so-called ``hidden'' barriers in the free energy landscape, often linked to slow conformational changes of a ligand or receptor.\cite{slow-barriers}.  If the slow degrees of freedom can be identified, then appropriate enhanced sampling schemes can be employed.  But because such bottlenecks are highly system-dependent,  it is difficult to overcome sampling barriers of these types in a general way, and past efforts have addressed such problems on a case-by-case basis.\cite{case-by-case}

Another source of bottlenecks arises from poor Monte Carlo acceptance probabilities, and these can be addressed in a much more general way. Although such sampling barriers may be hard to predict ahead of time, they can easily be identified using short simulations, and steps can be taken to adjust the parameters of the expanded ensemble algorithm, or introduce more lambda intermediates.

Our goal in this study is to demonstrate, using a very simple system of 1D harmonic oscillators, with no hidden barriers, the general phenomenon that Monte Carlo bottlenecks in Wang-Landau expanded ensemble sampling arise from broad distributions $P(\Delta u_{ij})$, leading to hysteresis in $\lambda$-sampling.   Sampling bottlenecks can be alleviated by introducing more thermodynamic ensembles at intermediate values of $\lambda$, but not so many as to create additional travel time.

To develop some simple rules for how to choose the best algorithm parameters, we construct a simple continuous-time kinetic model of the evolution of free energy estimates, which predicts conditions under which hysteresis occurs.  \color{red} This model suggests that $\lambda$ intermediates be chosen to achieve at least $\sim XX\%$ acceptance probability between neighboring ensembles .... VAV:  I don't know what this will say yet. :) \color{black}

@@

\subsection{The Wang-Landau algorithm}

@@

\subsection*{A simple model of expanded ensemble sampling bottlenecks}

@@

Markov State Model approaches ...

\begin{acknowledgments}
The authors thank the participants of Folding@home, without whom this work would not be possible.  We thank D. E. Shaw Research for providing access to folding trajectory data. This research was supported in part by the National Science Foundation through major research instrumentation grant number CNS-09-58854 and National Institutes of Health grants 1R01GM123296 and NIH Research Resource Computer Cluster Grant S10-OD020095.
\end{acknowledgments}

@@

% \nocite{*}
\bibliography{main}% Produces the bibliography via BibTeX.

>>>>

\begin{abstract}
Expanded ensemble algorithms can sample multiple thermodynamic states in a single-replica simulation to estimate free energies, and provide easily interpretable results.  In the case of receptor-ligand binding simulations, this entails sampling a schedule of thermodynamic intermediates which discretely de-couple the non-bonded interactions of the ligand, requiring the implementation of some form of restraint on the ligand's degrees of freedom to maintain its approximate orientation in largely de-coupled ensembles. While one could implement a simple harmonic center-of-mass restraint or more rigid Boresch restraints to ameliorate this problem, the prior does not help to restrict ligand orientation, and the latter can be more complicated to implement in an automated fashion for a large number of systems. Here, we show that by triangulating the ligand with three position restraints, we can accurately account for the bias of restricting the molecular degrees of freedom and that the free energy estimated by an expanaded ensemble scheme matches our derived analytical correction term in both a Markov Chain toy model and all-atom molecular dynamics.
\end{abstract}

%\pacs{Valid PACS appear here}% PACS, the Physics and Astronomy
                             % Classification Scheme.
\keywords{alchemical free energy perturbation, restraints, ligand binding, Monte Carlo, molecular dynamics}%Use showkeys class option if keyword
                              %display desired
\maketitle

@@

\section*{Introduction}
% In the last few years, there has been great interest in using expanded-ensemble (EE) methods\cite{lyubartsev1992new} to perform absolute and relative ligand binding free energy calculations (CITE Shirts papers, others).\cite{monroe2014converging}

% In the recent SAMPL6 challenge to predict host-guest affinities, EE methods performed with similar accuracy and efficiency to Hamiltonian replica exchange (HREX), with the advantage that the calculation only requires a single simulation that does not have to be coupled to other replicas.\cite{rizzi2020sampl6}   This property is highly desirable for large-scale \textit{in silico} screening efforts performed asynchronously using massively parallel distributed computing platforms like Folding@home, as pursued recently by Hurley et al.

Expanded ensemble algorithms allow multiple thermodynamic ensembles to be sampled in a single simulation.   This may be particularly useful for large-scale virtual screening on distributed computing platforms.

EE methods perform well at predicting absolute binding free energies of ligands (in SAMPL challenges), and relative binding free energies for small-molecule inhbitors (Si Zhang et al).   But EE method have not seen widespread adoption, in part because of sampling problems that often arise in practice.

One type of problem commonly seen in EE approaches is the determination of an appropriate restraint protocol to implement. Restraining ligand degrees of freedom in such simulations is required to predict accurate free energies of binding since the ligand must sample the bound pose throughout the simulation and will unbind once the non-bonded interactions are de-coupled if no restraints are used.

Using the standard Wang-Landau flat-histogram algorithm as implemented in the GROMACS molecular dynamics simulation software, we determined that the bias induced from restraining ligand degrees of freedom must be accounted for, and preferably with a simple analytical correction term that may be used in post-processing.

In this work, we develop a simple model based on a position-restrained Lennard-Jones trimer and show that this protocol can be scaled up to larger ligands up to a certain number of rotatable bonds, at which point ligand flexibility and distances between the triangulated position restraints is insufficient to properly restrain the ligand orientation.

@@

%In practice, bottlenecks often arise in the Monte Carlo sampling that lead to poor convergence of the free energy estimates.  One source of sampling bottlenecks comes from so-called ``hidden'' barriers in the free energy landscape, often linked to slow conformational changes of a ligand or receptor.\cite{slow-barriers}.  If the slow degrees of freedom can be identified, then appropriate enhanced sampling schemes can be employed.  But because such bottlenecks are highly system-dependent,  it is difficult to overcome sampling barriers of these types in a general way, and past efforts have addressed such problems on a case-by-case basis.\cite{case-by-case}

%Another source of bottlenecks arises from poor Monte Carlo acceptance probabilities, and these can be addressed in a much more general way. Although such sampling barriers may be hard to predict ahead of time, they can easily be identified using short simulations, and steps can be taken to adjust the parameters of the expanded ensemble algorithm, or introduce more lambda intermediates.

%Our goal in this study is to demonstrate, using a very simple system of 1D harmonic oscillators, with no hidden barriers, the general phenomenon that Monte Carlo bottlenecks in Wang-Landau expanded ensemble sampling arise from broad distributions $P(\Delta u_{ij})$, leading to hysteresis in $\lambda$-sampling.   Sampling bottlenecks can be alleviated by introducing more thermodynamic ensembles at intermediate values of $\lambda$, but not so many as to create additional travel time.

%To develop some simple rules for how to choose the best algorithm parameters, we construct a simple continuous-time kinetic model of the evolution of free energy estimates, which predicts conditions under which hysteresis occurs.  \color{red} This model suggests that $\lambda$ intermediates be chosen to achieve at least $\sim XX\%$ acceptance probability between neighboring ensembles .... VAV:  I don't know what this will say yet. :) \color{black}

@@

\subsection{A Lennard-Jones Toy Model}

\subsection{All-Atom Molecular Dynamics}

@@

\subsection*{A simple model for describing position restraints for estimating binding free energies}

@@

\begin{acknowledgments}
This research was supported in part by the National Science Foundation through major research instrumentation grant number CNS-09-58854 and National Institutes of Health grants 1R01GM123296 and NIH Research Resource Computer Cluster Grant S10-OD020095.
\end{acknowledgments}

@@

% \nocite{*}
\bibliography{toy-ee}% Produces the bibliography via BibTeX.
